\textbf{Spatio-temporal distinct patterns in variations of \(PM_{10}\) and \(PM_{2.5}\) relative to the recent drivings of emission sources in Mongolia}

### 2024-11-19 submission/manuscript.Rmd
Drafting Results

Mongolian dust is well-known by its term of coarse, \* color of brown,
\* its influence on downwind regions, \* and big role in global dust, \*
and its seasonality.

Because, it mainly consists of coarse fractions researchers neglect its
role in global climate into elaborating global dust-aerosol effects.
*However, recent change in the driving of the emissions of particulate
matters might cause fluctuations* its spatio-temporal variations and \*
characteristics and \* emerge a new pattern, eventually it may lead to
the role in the regional and global dust, so on climate system through
\* altering solar incidence, cloud formation, and precipitation or
warming \* the climate with the fraction of finer particulates.

Mongolia might turn into the not only coarse-dust source region but also
fine-mixed coarse-dust region.

-   Use of coal is increased

-   Urbanization

-   The monthly mean concentrations of PM10 (PM2.5) reached annual
    maximum in December and January due to winter synoptic governing
    conditions in Ulaanbaatar, capital city of Mongolia (Jugder).

-   Despite this, the spring dust storms creates another polluted season
    in UB. On spring, the dust storm from the Gobi Desert contribute
    significantly to increased aerosols in the atmosphere and ambient
    air pollution, leading to sporadic peaks in PM10 concentrations
    reaching as high as 64-234 $\mu g m^{-3}$ per day or exceeding 6000
    $\mu g m^{-3}$ per hour (Jugder).

-   Siberian anticyclonic activity governed over Mongolia, which create
    a significant vulnerability to winter air pollution in the populated
    areas.

-   A such changes in PM10 and PM2.5 to stagnant weather conditions, and
    local or transported dust was also observed in other countries China
    (Wang), Korea (Kim) and Japan ().

-   deposition

-   concentrations of particulate matter is ephederemal, yet vary
    depending on whether the pollution cause is natural or industrial,
    local or transported, seasonal or non-seasonal, makes complex and
    challenging.

Therefore, we aimed to demonstrate the distinct temporal and spatial
variations of PM2.5 and PM10 across urban and rural Mongolia using
extensive data from 2008 to 2020.

-   The present study will contribute significantly to the understanding
    of air particulate matter patterns in Mongolia and providing
    comprehensive data insights for policymakers and public health
    sectors.
-   Our findings is useful not only for addressing national health
    impacts but also beneficial for understanding air particulate matter
    as ambient air pollution, and tackling atmospheric aerosol effects
    in the climate system, and revealing their transboundary effects to
    the downwind regions in South-east Asia.

>>>>

## Research Qs
Therefore, we aimed to demonstrate the distinct temporal and spatial
variations of PM2.5 and PM10 across urban and rural Mongolia using
extensive data from 2008 to 2020.

On spring, the dust storm from the Gobi Desert contribute
significantly to increased aerosols in the atmosphere and ambient
air pollution, leading to sporadic peaks in PM10 concentrations
reaching as high as 64-234 $\mu g m^{-3}$ per day or exceeding 6000
$\mu g m^{-3}$ per hour (Jugder).
concentrations of particulate matter is ephederemal, yet vary
    depending on whether the pollution cause is natural or industrial,
    local or transported, seasonal or non-seasonal, makes complex and
    challenging.
    1.  Do concentrations of particulate matters differ in between urban
        and rural sites, and even within Gobi sites?
    2.  Do distinct temporal variations has existed among the sites?
    3.  Do PM2.5 particulates has contributed to the PM10 annual
        variations?

    -   If yes, how much, and when and where?
    -   What is the sd, mean, and median
        -   box plot
        -   violin
        -   scatter points, epidemic, sporadic
    -   Daily variations to examine it related to the heating
        -   2 peaks: smaller and bigger
        -   compare the t-duration exceeds 50mug/m3/hour

    4.  Does it has distinct patterns among the sites regarding to the
        drivings

    -   How PMs varies with the wind speed and visibility
    -   Do they differently explained with variables and changes in
        drivings (with PCA analysis)

    5.  Is there any significant changes in time-series of PMs at 4
        seasons
    6.  Is there any significant changes in ratio in the spring in
        respect to winter?

        The present study will contribute significantly to the understanding
            of air particulate matter patterns in Mongolia and providing
            comprehensive data insights for policymakers and public health
            sectors. Our findings is useful not only for addressing national health
            impacts but also beneficial for understanding air particulate matter
            as ambient air pollution, and tackling atmospheric aerosol effects
            in the climate system, and revealing their transboundary effects to
            the downwind regions in South-east Asia.

## Results
### The spatio-temporal variations of the PMs at the study sites
To comparatively examine the spatio-difference regarding to the emissions of PMs, we illustrated the hourly observed values of PM10 and PM2.5 for all sites (figure_3). On each site, PM10 is more sporadic than PM2.5 due to dust epidemic nature. In conjunction, the mean values averaged from hourly values are larger than its median concentrations both on PM10 and PM2.5 for all sites. The mean variables of p-value show that concentrations of particulate matters significantly differ with 99% confident levels at all sites (figure_3), except  95% confident level between DZ and UB on PM10 (figure_3a). It clearly exhibits general variations of particulate matters as expected and emphasizes that the significant difference in the concentration values among the sites. Besides all sites are quantitatively differ by the values of concentrations in particulate matters, there are two main characteristics can be seen when we compare the median deviations from its mean values etc.

- such patterns are strongly manifested in ZU and SS sites.  
- the mean values those averaged from hourly concentrations of PM10 and PM2.5 is larger than its median are strongly manifested for all in UB and DZ sites.
- Additionally, consider the significance of the ..

This highlight that the fine and coarse particulate matters significantly vary among sites of urban and rural sites, and even within Gobi (rural) sites.
Moreover, .... has a diversely characteristically diversified in respect with the PM2.5 (fine) to PM10 (coarse) particulates.,. which requires urban impacts

To reveal the natural and anthropogenic impacts on the concentrations of particulate matters, we demonstrated annual variations of PM10 and PM2.5 for each sites (figure_4).
3.  Do PM2.5 particulates has contributed to the PM10 annual
    variations?
      Distinct temporal variations has existed among the sites.
        -   PM2.5 particulates has contributed to the PM10 annual
            variations in UB and in DZ in winter.
            -   If yes, how much, and when and where?
            -   What is the sd, mean, and median
                -   box plot
                -   violin
                -   scatter points, epidemic, sporadic
            -   Daily variations to examine it related to the heating
                -   2 peaks: smaller and bigger
                -   compare the t-duration exceeds 50mug/m3/hour

[Therefore] SS and ZU sites are mainly affected by the spring dust, followed by the autumn dust. Annual maximum in the winter for DZ and UB are from PM2.5, which results an increase in PM10.
It requires the cause the behind such the variations. DZ site is polluted in the winter by the heating and in the spring by the natural dust.

[Therefore] Spatio-temporally in two class; consists of 2 Gobi sites, and 1 urban plus urbanized Gobi sites.

### The emission patterns of interrelations among meteorological variables at the study sites

        -   PMs varies with the wind speed and visibility
        -   In general, three distinct patterns were resulted with PCA
            analysis, which is in consistent with temporal variation.
            explained with variables and changes in drivings (with PCA
            analysis)

DZ site is polluted in the winter by the heating and in the spring by the natural dust.

### The recent trends in concentrations of PMS and fine-coarse

        fractional changes at the sites
        -   There are significant changes in time-series of PMs at 4
            seasons
        -   There any significant changes in ratio in the spring in
            respect to winter in DZ.
        -   Close relationships was found between PM2.5 in winter and r
            values in the spring.

-   The spatio-temporal variations of the PMs at the study sites
        -   Concentrations of particulate matters differ in between
            urban and rural sites, and even within Gobi sites.
        -   Distinct temporal variations has existed among the sites.
        -   PM2.5 particulates has contributed to the PM10 annual
            variations in UB and in DZ in winter.
            -   If yes, how much, and when and where?
            -   What is the sd, mean, and median
                -   box plot
                -   violin
                -   scatter points, epidemic, sporadic
            -   Daily variations to examine it related to the heating
                -   2 peaks: smaller and bigger
                -   compare the t-duration exceeds 50mug/m3/hour
    -   The emission patterns of interrelations among meteorological
        variables at the study sites
        -   PMs varies with the wind speed and visibility
        -   In general, three distinct patterns were resulted with PCA
            analysis, which is in consistent with temporal variation.
            explained with variables and changes in drivings (with PCA
            analysis)
    -   The recent trends in concentrations of PMS and fine-coarse
        fractional changes at the sites
        -   There are significant changes in time-series of PMs at 4
            seasons
        -   There any significant changes in ratio in the spring in
            respect to winter PM2.5 in DZ.
    -   Close relationships was found between PM2.5 in winter and r
        values in the spring. Thus, our research results clearly proves
        the distinct variations in PMs has emerged. The dust fine-coarse
        fractions was manifested at the town center for the Gobi sites,
        which reveals the that Mongolian dust composites not only coarse
        dust, but also fine particulate matters. The particulates likely
        consisted of the black carbon, which may give a substantial
        effect on climate systems. if this trend continues on as coal
        consumption with the population growth in the future.
    -   CO Carbon monoxide is obtained due to incomplete combustion of
        charcoal in a closed room.
    -   CO2

### 2024-11-19 submission/manuscript.tex
Drafting Results

Mongolian dust is well-known by its term of coarse, * color of brown, *
its influence on downwind regions, * and big role in global dust, * and
its seasonality.

Because, it mainly consists of coarse fractions researchers neglect its
role in global climate into elaborating global dust-aerosol effects.
\emph{However, recent change in the driving of the emissions of
particulate matters might cause fluctuations} its spatio-temporal
variations and * characteristics and * emerge a new pattern, eventually
it may lead to the role in the regional and global dust, so on climate
system through * altering solar incidence, cloud formation, and
precipitation or warming * the climate with the fraction of finer
particulates.

Mongolia might turn into the not only coarse-dust source region but also
fine-mixed coarse-dust region.

\begin{itemize}
\item
  Use of coal is increased
\item
  Urbanization
\item
  The monthly mean concentrations of PM10 (PM2.5) reached annual maximum
  in December and January due to winter synoptic governing conditions in
  Ulaanbaatar, capital city of Mongolia (Jugder).
\item
  Despite this, the spring dust storms creates another polluted season
  in UB. On spring, the dust storm from the Gobi Desert contribute
  significantly to increased aerosols in the atmosphere and ambient air
  pollution, leading to sporadic peaks in PM10 concentrations reaching
  as high as 64-234 \(\mu g m^{-3}\) per day or exceeding 6000
  \(\mu g m^{-3}\) per hour (Jugder).
\item
  Siberian anticyclonic activity governed over Mongolia, which create a
  significant vulnerability to winter air pollution in the populated
  areas.
\item
  A such changes in PM10 and PM2.5 to stagnant weather conditions, and
  local or transported dust was also observed in other countries China
  (Wang), Korea (Kim) and Japan ().
\item
  deposition
\item
  concentrations of particulate matter is ephederemal, yet vary
  depending on whether the pollution cause is natural or industrial,
  local or transported, seasonal or non-seasonal, makes complex and
  challenging.
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  The present study will contribute significantly to the understanding
  of air particulate matter patterns in Mongolia and providing
  comprehensive data insights for policymakers and public health
  sectors.
\item
  Our findings is useful not only for addressing national health impacts
  but also beneficial for understanding air particulate matter as
  ambient air pollution, and tackling atmospheric aerosol effects in the
  climate system, and revealing their transboundary effects to the
  downwind regions in South-east Asia.
\end{itemize}

@@

>>>>

\subsection{Research Qs}\label{research-qs}

@@

On spring, the dust storm from the Gobi Desert contribute significantly
to increased aerosols in the atmosphere and ambient air pollution,
leading to sporadic peaks in PM10 concentrations reaching as high as
64-234 \(\mu g m^{-3}\) per day or exceeding 6000 \(\mu g m^{-3}\) per
hour (Jugder). concentrations of particulate matter is ephederemal, yet
vary depending on whether the pollution cause is natural or industrial,
local or transported, seasonal or non-seasonal, makes complex and
challenging. 1. Do concentrations of particulate matters differ in
between urban and rural sites, and even within Gobi sites? 2. Do
distinct temporal variations has existed among the sites? 3. Do PM2.5
particulates has contributed to the PM10 annual variations?

\begin{verbatim}
-   If yes, how much, and when and where?
-   What is the sd, mean, and median
    -   box plot
    -   violin
    -   scatter points, epidemic, sporadic
-   Daily variations to examine it related to the heating
    -   2 peaks: smaller and bigger
    -   compare the t-duration exceeds 50mug/m3/hour

4.  Does it has distinct patterns among the sites regarding to the
    drivings

-   How PMs varies with the wind speed and visibility
-   Do they differently explained with variables and changes in
    drivings (with PCA analysis)

5.  Is there any significant changes in time-series of PMs at 4
    seasons
6.  Is there any significant changes in ratio in the spring in
    respect to winter?

    The present study will contribute significantly to the understanding
        of air particulate matter patterns in Mongolia and providing
        comprehensive data insights for policymakers and public health
        sectors. Our findings is useful not only for addressing national health
        impacts but also beneficial for understanding air particulate matter
        as ambient air pollution, and tackling atmospheric aerosol effects
        in the climate system, and revealing their transboundary effects to
        the downwind regions in South-east Asia.
\end{verbatim}

\subsection{Results}\label{results}

\subsubsection{The spatio-temporal variations of the PMs at the study
sites}\label{the-spatio-temporal-variations-of-the-pms-at-the-study-sites}

To comparatively examine the spatio-difference regarding to the
emissions of PMs, we illustrated the hourly observed values of PM10 and
PM2.5 for all sites (figure\_3). On each site, PM10 is more sporadic
than PM2.5 due to dust epidemic nature. In conjunction, the mean values
averaged from hourly values are larger than its median concentrations
both on PM10 and PM2.5 for all sites. The mean variables of p-value show
that concentrations of particulate matters significantly differ with
99\% confident levels at all sites (figure\_3), except 95\% confident
level between DZ and UB on PM10 (figure\_3a). It clearly exhibits
general variations of particulate matters as expected and emphasizes
that the significant difference in the concentration values among the
sites. Besides all sites are quantitatively differ by the values of
concentrations in particulate matters, there are two main
characteristics can be seen when we compare the median deviations from
its mean values etc.

\begin{itemize}
\tightlist
\item
  such patterns are strongly manifested in ZU and SS sites.\\
\item
  the mean values those averaged from hourly concentrations of PM10 and
  PM2.5 is larger than its median are strongly manifested for all in UB
  and DZ sites.
\item
  Additionally, consider the significance of the ..
\end{itemize}

This highlight that the fine and coarse particulate matters
significantly vary among sites of urban and rural sites, and even within
Gobi (rural) sites. Moreover, \ldots. has a diversely characteristically
diversified in respect with the PM2.5 (fine) to PM10 (coarse)
particulates.,. which requires urban impacts

To reveal the natural and anthropogenic impacts on the concentrations of
particulate matters, we demonstrated annual variations of PM10 and PM2.5
for each sites (figure\_4). 3. Do PM2.5 particulates has contributed to
the PM10 annual variations? Distinct temporal variations has existed
among the sites. - PM2.5 particulates has contributed to the PM10 annual
variations in UB and in DZ in winter. - If yes, how much, and when and
where? - What is the sd, mean, and median - box plot - violin - scatter
points, epidemic, sporadic - Daily variations to examine it related to
the heating - 2 peaks: smaller and bigger - compare the t-duration
exceeds 50mug/m3/hour

{[}Therefore{]} SS and ZU sites are mainly affected by the spring dust,
followed by the autumn dust. Annual maximum in the winter for DZ and UB
are from PM2.5, which results an increase in PM10. It requires the cause
the behind such the variations. DZ site is polluted in the winter by the
heating and in the spring by the natural dust.

{[}Therefore{]} Spatio-temporally in two class; consists of 2 Gobi
sites, and 1 urban plus urbanized Gobi sites.

\subsubsection{The emission patterns of interrelations among
meteorological variables at the study
sites}\label{the-emission-patterns-of-interrelations-among-meteorological-variables-at-the-study-sites}

\begin{verbatim}
    -   PMs varies with the wind speed and visibility
    -   In general, three distinct patterns were resulted with PCA
        analysis, which is in consistent with temporal variation.
        explained with variables and changes in drivings (with PCA
        analysis)
\end{verbatim}

DZ site is polluted in the winter by the heating and in the spring by
the natural dust.

\subsubsection{The recent trends in concentrations of PMS and
fine-coarse}\label{the-recent-trends-in-concentrations-of-pms-and-fine-coarse}

\begin{verbatim}
    fractional changes at the sites
    -   There are significant changes in time-series of PMs at 4
        seasons
    -   There any significant changes in ratio in the spring in
        respect to winter in DZ.
    -   Close relationships was found between PM2.5 in winter and r
        values in the spring.
\end{verbatim}

\begin{itemize}
\tightlist
\item
  The spatio-temporal variations of the PMs at the study sites -
  Concentrations of particulate matters differ in between urban and
  rural sites, and even within Gobi sites. - Distinct temporal
  variations has existed among the sites. - PM2.5 particulates has
  contributed to the PM10 annual variations in UB and in DZ in winter. -
  If yes, how much, and when and where? - What is the sd, mean, and
  median - box plot - violin - scatter points, epidemic, sporadic -
  Daily variations to examine it related to the heating - 2 peaks:
  smaller and bigger - compare the t-duration exceeds 50mug/m3/hour

  \begin{itemize}
  \tightlist
  \item
    The emission patterns of interrelations among meteorological
    variables at the study sites

    \begin{itemize}
    \tightlist
    \item
      PMs varies with the wind speed and visibility
    \item
      In general, three distinct patterns were resulted with PCA
      analysis, which is in consistent with temporal variation.
      explained with variables and changes in drivings (with PCA
      analysis)
    \end{itemize}
  \item
    The recent trends in concentrations of PMS and fine-coarse
    fractional changes at the sites

    \begin{itemize}
    \tightlist
    \item
      There are significant changes in time-series of PMs at 4 seasons
    \item
      There any significant changes in ratio in the spring in respect to
      winter PM2.5 in DZ.
    \end{itemize}
  \item
    Close relationships was found between PM2.5 in winter and r values
    in the spring. Thus, our research results clearly proves the
    distinct variations in PMs has emerged. The dust fine-coarse
    fractions was manifested at the town center for the Gobi sites,
    which reveals the that Mongolian dust composites not only coarse
    dust, but also fine particulate matters. The particulates likely
    consisted of the black carbon, which may give a substantial effect
    on climate systems. if this trend continues on as coal consumption
    with the population growth in the future.
  \item
    CO Carbon monoxide is obtained due to incomplete combustion of
    charcoal in a closed room.
  \item
    CO2
  \end{itemize}
\end{itemize}

@@

\subsection{Conclusions}\label{conclusions-1}
